### 2025-09-25 paper/sec/evaluation.tex
Tweak.

To what extent can LLMs accurately interpret quantitative and semi-quantitative claims in scholarly text as
data queries? We examine performance across a range of linguistic idioms (e.g. averages, percentages, min/max,
ranks, as summarised in Table~\ref{tab:natural-language-forms}) and investigate how accuracy varies with task
complexity, measured (somewhat crudely and approximately) by the number of query sub-expressions (e.g.
retrieval, aggregation, or arithmetic) present in the gold solution.

@@

How well do the generated expressions generalise when the underlying data changes, or when the input contains
misleading or ill-specified information? We test whether generated queries continue to produce correct outputs
under a set of hand-generated counterfactual modifications of the dataset, based on expected query results
specific to each test case, and also how counterfactual performance is impacted by the presence of misleading
or adversarial phrasing. able~\ref{tab:problematic_cases} shows some of cases which are ``problematic'' in
this sense; in these cases, producing a valid expression is likely to be challenging because of ambiguities in
the input data or accompanying natural language.

@@

To evaluate RQ1, we used a sample of the \SciGen dataset~\citep{moosavi21}, an open source dataset consisting
of tables from scientific articles and their corresponding descriptions. We aggregated the generated
expression in the linguistic categories from Table~\ref{tab:natural-language-forms}.
Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and
without target-value sharing.

@@

\subsection{Counterfactual testing}
\label{sec:validation-with-counterfactual-data}

For the \emph{counterfactual testing} part of our analysis, the input data is deliberately modified in order
to verify the consistency of the behaviours. Both the expected and generated expressions are executed again
and the results are compared. If discrepancies are found the \InterpretationAgent ask for a new generation
sending back that the failure is related to a counterfactual error.

%\rpnote{complete}

Across 300 test executions, 121 contained at least one counterfactual error (avg. 3.8), of which 42 ultimately
succeeded. Such tests help detect expressions that may yield the expected output but not the correct solution.
For example, the \InterpretationAgent generated:

\begin{verbatim}
(findWithKey_ "model" "LSTM2" tableData).time_s
\end{verbatim}

intended to retrieve the execution time of the \texttt{LSTM} model, but incorrectly referred to \texttt{LSTM2}.
Counterfactual testing exposed this error and prompted regeneration.

>>>>

To what extent can LLMs accurately interpret quantitative and semi-quantitative claims in scholarly text as
data queries? We examine performance across a range of linguistic idioms (e.g. averages, percentages, min/max,
ranks, as summarised in Table~\ref{tab:natural-language-forms}) and investigate how accuracy varies with task
complexity, measured (somewhat crudely) by the number of query sub-expressions (e.g. retrieval, aggregation,
or arithmetic) present in the gold solution.

@@

How well do the generated expressions generalise when the underlying data changes, or when the input contains
misleading or ill-specified information? We test whether generated queries continue to produce correct outputs
under a set of hand-generated counterfactual modifications of the dataset, based on expected query results
specific to each test case, and also how counterfactual performance is impacted by the presence of misleading
or adversarial phrasing. Table~\ref{tab:problematic_cases} shows some of cases we deem problematic in this
sense; in these cases, producing a valid expression is likely to be challenging because of ambiguities in the
input data or accompanying natural language.

@@

\paragraph{Interpretation Accuracy across Linguistic Idioms and Complexity.} To evaluate RQ1, we used a sample
of the \SciGen dataset~\citep{moosavi21}, an open source dataset consisting of tables from scientific articles
and their corresponding descriptions. We aggregated the results according to the linguistic categories from
Table~\ref{tab:natural-language-forms}. Figure~\ref{fig:success_rate_by_category} illustrates the success rate
for each category, both with and without target-value sharing.

@@

\paragraph{Generalisability and Robustness.} As a preliminary attempt to address RQ2, we carried out
\emph{counterfactual testing} to evaluate the robustness of generated expressions under changes to the
underlying data. In this setup, the input tables were modified according to hand-craft test specifications,
and both the expected and generated expressions were re-executed to check whether the behaviours remained
consistent. Across 300 test executions, 121 contained at least one counterfactual error (an average of 3.8 per
case), of which 42 ultimately still succeeded. These tests highlight cases where an expression may
coincidentally yield the correct output on the original data but fails to be extensionally equivalent more
generally (i.e.~under perturbation). For example, in one test the system generated

\hspace{3mm} \kw{(findWithKey\_ "model" "LSTM2" tableData).time\_s}

intended to retrieve the execution time of the \texttt{LSTM} model, but incorrectly referred to \texttt{LSTM2}.
Counterfactual testing exposed this mismatch, which would otherwise have gone undetected.

At present, counterfactual tests are used only as an evaluation device, not as part of the authoring workflow
itself. For future work (\secref{conclusion}), we plan to investigate automatic generation of counterfactual
tests, allowing these additional robustness checks to be integrated into the document authoring workflow.
